An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-22 comment.tex
ttt

\section{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the the
throughput of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30,
  BayesianDorfman, MatrixPooling, LionDorfman}. These techniques,
often termed \emph{pooling}, or group testing, have originated in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling for has been demonstrated empirically \cite{DorfmanYuvalDor}
and its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

@@

\subsection{Refuting common assumption}\label{subsec:refute}
We refute the common assumption with experimental data summarized in
Table \ref{table}, collected from \cite{Salazar}. There, the authors
investigate Dorfman pooling and, regardless of the pooled test result,
follow up and test each pool member separately. We focus on 128 pools
for which at least one subsequent separate test was positive --- of
which 29 pooled tests were negative and 99 positive. In the data cited
in \cite{Salazar}, of the 29 negative pools, subsequent separate
testing yielded a single positive result in 24. In contrast, of the 99
positive pools, 42 yielded a single positive test upon subsequent
separate testing.

@@

To summarize, we assume that for a single pool, a positive test result
is generated in one of two paths. Either, SARS-CoV-2 RNA from an
infected individual's sample is correctly amplified and detected, and
this can happen for each positive sample in a pool, with probability
$\Se$. Or, some erroneous amplification occurs (e.g. contaminant viral
RNA is introduced), an event that occurs at most once per pool, with
probability $1-\Sp$. Our model is illustrated in
Figure~\ref{fig:likelihood}, and summarized in
%% A negative pooled test result occurs when all detection paths (both
%% correct and erroneous) fail. The probability of no false-detection
%% accounts for the $\spec$ term. The probability of no correct detection
%% is $(1-\sens)$ per infected individual. The probability that all such
%% paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

@@

\subsection{Application}
As an application of our model, we calculate the differnece in
false-negative rate for a single \emph{infected} individual,
henceforth referred to as "Donald". We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our goal is to calculate Dorfman's scheme
false-negative rate. Rephrasing, we wish to answer the following
question: what is the probability of not identifying Donald as
infected under a Dorfman pooling scheme?

\subsubsection{Scheme false-negative rate}
Denote the prevalence of infection in the (tested) population $q$. We
give Donald the respect he deserves and assign him as individual $1$,
so that $\theta_1=1$. Then:

@@

Our results have practical implications for the use of
pooling schemes. As the COVID-19 pandemic progresses, the infection prevalence in various tested populations undergoes frequent changes. Hence, as our results suggest, pooling strategies for mass testing should be used with caution in populations where infection rates are low. Such mass-tested populations often include air travel passengers, potentially importing infections, or asymptomatic individuals, such as children,  and can be crucial for controlling outbreaks(ref)

>>>>

\section{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the the
throughput of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30,
  BayesianDorfman, MatrixPooling, LionDorfman}. These techniques,
often termed \emph{pooling}, or group testing, have originated in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor}
and its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

Studies focused on pooling for SARS-CoV-2 are in a consensus that sample dilution 
effects \cite{Dilution1, Dilution2} are not a concern \cite{Lion, DorfmanYuvalDor}.

@@

\subsection{Refuting the common assumption}\label{subsec:refute}
We refute the common assumption with experimental data summarized in
Table \ref{table}, collected from \cite{Salazar}. There, the authors
investigate Dorfman pooling and, regardless of the pooled test result,
follow up and test each pool member separately. We focus on 128 pools
for which at least one subsequent separate test was positive --- of
which 29 pooled tests were negative and 99 positive. In the data cited
in \cite{Salazar}, of the 29 negative pools, subsequent separate
testing yielded a single positive result in 24. In contrast, of the 99
positive pools, 42 yielded a single positive test upon subsequent
separate testing.

@@

To summarize, we assume that for a single pool, a positive test result
is generated in one of two paths. Either, SARS-CoV-2 RNA from an
infected individual's sample is correctly amplified and detected, and
this can happen for each positive sample in a pool, with probability
$\Se$. Or, some erroneous amplification occurs (e.g. contaminant viral
RNA is introduced), an event that occurs at most once per pool, with
probability $1-\Sp$. Our model is illustrated in
Figure~\ref{fig:likelihood} and summarized in
%% A negative pooled test result occurs when all detection paths (both
%% correct and erroneous) fail. The probability of no false-detection
%% accounts for the $\spec$ term. The probability of no correct detection
%% is $(1-\sens)$ per infected individual. The probability that all such
%% paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

@@

\subsection{Application}
As an application, we calculate the false-negative rate for a single \emph{infected} individual,
henceforth referred to as "Donald" under our model and under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Rephrasing, we wish to answer the following
question: what is the probability of not identifying Donald as
infected under a Dorfman pooling scheme?

\subsubsection{Scheme false-negative rate}
Denote the prevalence of infection in the (tested) population $q$. We
assign Donald as individual $1$, so that $\theta_1=1$. Then:

@@

Our results have practical implications for the use of
pooling schemes. As the COVID-19 pandemic progresses, the infection prevalence in various tested populations undergoes frequent changes. Hence, as our results suggest, pooling strategies for mass testing should be used with caution in populations where infection rates are low. Such mass-tested populations often include air travel passengers \cite{JTM}, potentially importing infections, or asymptomatic individuals, such as children, and can be crucial for controlling outbreaks \cite{MinaScience}.
